---
title: "Missing Data"
output: 
  slidy_presentation:
    css: styles5293.css
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      cache = TRUE, message = FALSE)
```


# Looking for patterns in missing data

## row / column patterns

> - Do all variables have the same percentage of values missing? 

> - Are there row patterns to the missing values?

# Create a dataset with missing values

```{r}
# Add NAs to mtcars dataset
set.seed(5702)
mycars <- mtcars
mycars[,"gear"] <- NA
mycars[10:20, 3:5] <- NA
for (i in 1:10) mycars[sample(32,1), sample(11,1)] <- NA
```

# Row / column missing patterns

## Missing values by column

```{r, echo = TRUE}
colSums(is.na(mycars)) %>%
  sort(decreasing = TRUE)
```

# Row / column missing patterns

## Missing values by row

```{r, echo = TRUE}
rowSums(is.na(mycars)) %>%
  sort(decreasing = TRUE)
```


# Row / column missing patterns

## heatmap

`geom_tile()`

```{r}
library(tidyverse)

tidycars <- mycars %>% 
    rownames_to_column("id") %>% 
    gather(key, value, -id) %>% 
    mutate(missing = ifelse(is.na(value), "yes", "no"))
```

```{r}
ggplot(tidycars, aes(x = key, y = fct_rev(id), fill = missing)) +
  geom_tile(color = "white") + 
  ggtitle("mtcars with NAs added") +
  scale_fill_viridis_d() + # discrete scale
  theme_bw()
```


# Row / column missing patterns

## heatmap

`mi::missing_data.frame()` 

```{r, echo = TRUE, results = FALSE}
library(mi)
x <- missing_data.frame(mycars)
```

```{r, echo = TRUE}
image(x)
```


(gear not shown since all are missing)

# Row / column missing patterns

## Missing values by variable

`geom_tile()`

```{r}
ggplot(tidycars, aes(x = key, y = fct_rev(id), fill = value)) +
  geom_tile(color = "white") + 
  scale_fill_gradient(low = "grey80", high = "red", na.value = "black") + theme_bw()
```

# Row / column missing patterns

## Missing values by variable

`geom_tile()` with standardized variables

```{r}
tidycars <- tidycars %>% group_by(key) %>% 
  mutate(Std = (value-mean(value, na.rm = TRUE))/sd(value, na.rm = TRUE)) %>% ungroup()

ggplot(tidycars, aes(x = key, y = fct_rev(id), fill = Std)) +
  geom_tile(color = "white") + 
  scale_fill_gradient2(low = "blue", mid = "white", high ="yellow", na.value = "black") + theme_bw()
```

# Row / column missing patterns

## Missing values by variable

reordered by number of missing


```{r}
# convert missing to numeric so it can be summed up
tidycars <- tidycars %>% 
  mutate(missing2 = ifelse(missing == "yes", 1, 0))

ggplot(tidycars, aes(x = fct_reorder(key, -missing2, sum), y = fct_reorder(id, -missing2, sum), fill = Std)) +
  geom_tile(color = "white") + 
  scale_fill_gradient2(low = "blue", mid = "white", high ="yellow", na.value = "black") + theme_bw()
```


# Row / column missing patterns

### Missing values by variable

<img src = "missingpatternskey.png" width = "600"></img>

# Row / column missing patterns

### Missing values by variable

missing patterns

```{r, echo = TRUE}
x <- mi::missing_data.frame(mycars)
class(x)
x@patterns
levels(x@patterns)
summary(x@patterns)
```

# Row / column missing patterns

### Missing values by variable

missing patterns

(repeated patterns are reduced to one row)

```{r, fig.height = 3}
library(extracat)
visna(mycars)
```

# visna() w/ row sort

Sorted by most common to least common missing *pattern* (top to bottom)

```{r, fig.height = 3}
visna(mycars, sort = "r")
```

# visna() w/ column sort

Sorted by *variable* with the most to least missing values (left to right)

```{r, fig.height = 3}
visna(mycars, sort = "c")
```

# visna() w/ row & column sort


```{r, fig.height = 3}
visna(mycars, sort = "b")
```


# NYC School data
```{r}
df <- read.csv("SAT2010.csv", na.strings = "s",
               check.names = FALSE)
head(df)
dim(df)
```


# NYC School data
```{r, fig.height = 3}
visna(df)
```

# Looking for patterns in missing data

## value patterns

> - Are missing patterns correlated with values of another variable?

> - Are certain value ranges more likely to be missing? 

> - Or are there no patterns at all? 

# NYC School data

Does the proportion of schools with missing data vary by borough?

Data: `SAT2010.csv`

# NYC School data

Does the proportion of schools with missing data vary by borough?

Data: `SAT2010.csv`


```{r}
head(df)
```

# Missing by borough

```{r}
df <- df %>% mutate(Borough = str_sub(DBN, 3, 3))
percent_missing <- df %>% group_by(Borough) %>% 
  summarize(num_schools = n(), num_na = sum(is.na(`Writing Mean`))) %>% 
  mutate(percent_na = round(num_na/num_schools, 2)) %>% 
  arrange(-percent_na)
percent_missing
```


# Missing by borough

K, Q, M, X, R

Manhattan

Brooklyn

Queens

The Bronx

Staten Island

# Missing by borough

K, Q, M, X, R

Manhattan = New York County

Brooklyn  = Kings County

Queens = Queens County

The Bronx = Bronx County

Staten Island = Richmond County


# Missing by borough

K, Q, M, X, R

<span style="color:red">M</span>anhattan = New York County = "M"

Brooklyn = <span style="color:red">K</span>ings County = "K"

<span style="color:red">Q</span>ueens = <span style="color:red">Q</span>ueens County = "Q"

The Bron<span style="color:red">x</span> = Bron<span style="color:red">x</span> County = "X"

Staten Island = <span style="color:red">R</span>ichmond County = "R"

# Missing by borough

```{r}
percent_missing <- percent_missing %>%
  mutate(BoroughName = fct_recode(Borough,
                                      Manhattan = "M",
                                      Brooklyn = "K", 
                                      Queens = "Q", 
                                      `The Bronx` = "X",
                                      `Staten Island` = "R"))
percent_missing

```

# Missing by borough

```{r}
df <- df %>% 
  mutate(BoroughName = fct_recode(Borough, 
                                      Manhattan = "M",
                                      Brooklyn = "K", 
                                      Queens = "Q", 
                                      `The Bronx` = "X",
                                      `Staten Island` = "R"))

dfsum <- df %>% group_by(BoroughName) %>% 
  summarize(Reading = round(mean(`Critical Reading Mean`, na.rm = TRUE), 1), 
            Math = round(mean(`Mathematics Mean`, na.rm = TRUE), 1),
            Writing = round(mean(`Writing Mean`, na.rm = TRUE), 1)) %>%
  left_join(percent_missing %>% select(BoroughName, percent_na), 
            by = "BoroughName") %>% 
  arrange(desc(percent_na))
```

# Missing by borough

```{r}
library(GGally)
dfsum
ggpairs(dfsum, columns = 2:5, aes(color = BoroughName))
```

